Stock forecasting algorithms and their usage

### 2015-06-08 Paper_self.tex
Added section about Lr, LDA, QDA

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
\usepackage[]{blindtext, graphicx}
\usepackage[hidelinks]{hyperref}
\graphicspath{ {Figures/} }
\usepackage{tabularx}
\usepackage{color}

@@

There is a huge variety of methods that are used in forecasting stock prices. Many of the algorithms that have been made-up just use structured data like tables with old stock prices in it. The aim of all these algorithms is to maximize the profit while holding the risk as low as possible.
\\
Since the stock market is global, there are many relationships between stocks. Some of them are paired with others so if the price of one stock goes down, the other stock will follow this trend.
Stocks are also influenced by numerous factors and these factors also have relationships between each other. This leads us to the thought, that a good stock prediction has to consider more than just old stock prices. However it is still possible to make a good prediction just based on old prices.
\\
One very important thing that has to be observed are the international newscasts. If the global economy, or even the economy of a single country is going down, the stock market will sooner or later follow that trend. But this trends are most likely to be relatively slow and easy to predict even looking at the stock prices itself. What is more dangerous are catastrophes. After a catastrophe there is often a low point on the stock market. This low point comes so fast, that only algorithms that observe the news have a chance to prevent huge losses.
\\
Another thing we want you to note is, that we did all our tests on old data of the stock market. Even if we \textit{bought} a stock at some specific point of time, we did not really buy this stock on the real market. So our tests did not affect the stock market at all. Because if you would really buy and sell stocks on the market while testing, it would always affect the results you get.

@@

In this paper we are going to compare some algorithms and their functionality. The main goal of these algorithms is, to predict a rise or fall in the price before the actual rise or fall happens. It is very important to find some kind of pattern in the stock prices of the past to do a accurate prediction of the future prices. We also like to share some of our experimental results we had with the algorithms. We only used algorithms that do not take global news in account because we wanted to see, how good the algorithms were in an isolated environment analysing past prices only. We decided to do it that way, because the algorithms themselves should work properly and examining news would be an easy addition to these working algorithms.
\\
The main goal of the project was, to find a good working stock forecast algorithm to predict future stock prices. Later in this paper, we will introduce parameters and variables, that give the best results combined with our algorithms.

\subsection{Document overview}
In Section \ref{essentials} we try to briefly cover the background information you need as the reader to understand the most important things about our project.
\\
Section \ref{algorithm} describes the algorithms we use in more detail. Every algorithm we used is self implemented. 
\\
Chapter \ref{results} is about the test-results we got using the algorithms described in chapter \ref{algorithm}. We tested all algorithms with different inputs and parameters.
\\
In the section \ref{conclusion} we will interpret the results and check how we can use our results maybe to improve our algorithms even more.
\\
In chapter \ref{futurework} we will give some thing our algorithms can be extended with and work that can be done in the future.

@@

We investigated algorithms that bring good results with time series analysis. The computer programs we implemented were specialized  making a good prediction only looking at one single stock and its trend. These algorithms were tested on real stock data we fetched from the yahoo finance API. 
\\
We used yahoo finance because it gives easy access to numerous stocks and their historical prices. The API yahoo grants is pretty simple. The only thing you have to do to download the data you need is, preparing a link so it fits your needs. For example the link: 
\\
\textit{\url{http://real-chart.finance.yahoo.com/table.csv?s=IBM&a=00&b=2&c=1962&d=04&e=21&f=2015&g=d&ignore=.csv}}
\\
 provides the data in a csv-file. The stock that this link is going to download is IBM (International Business Machines Corporation). It is also possible to adjust the timespan and format of the data. 
This method of getting historical stock market data is pretty easy to implement in various programming languages. It is also possible to download the csv-file to your computer with this link.
\\
After we got the data we needed, we started implementing the algorithms that we thought of.

% beschreiben wie der algorithmus von Herr Fischer funktioniert und die 2-3 die wir noch implementiert haben

@@

We implemented three different algorithms to predict stock prices and maximise the outcome. All three of them only consider one stock at a time and do not take any global news or trends in account. In this section we will describe the algorithms we used in detail

@@

With this algorithm we wanted to make a precise prediction on the stock market. It analyses the closing prices of a specific stock and gives a forecast how the price will develop. We also wanted the algorithm to be able to predict not only the price for the next day, but for the next few days. In section \ref{results} we will demonstrate that we were able to do a good prediction even 4 days in the future. But it must be said that the more in the future the prediction is, the bigger the gap between the real prices and the prediction.
\\
The algorithm itself uses only the last few days of a stock. The number of days used for this algorithm is not static. The amount of days that give the best result may vary for different stocks. Nevertheless if the number of days stays within a specific range, the results are pretty accurate.
\\
\\
\[ M = \sum_{n=0}^D C_{n+1} - C_{n} \]
\[ prediction = \frac{M}{D}\]
\[ stock_{predicted} = stock_{today} + prediction\]
\\
With D the number of days the prediction is based on and C the closing price of the nth day. After you calculated the prediction, the only thing that is left to do, is adding the prediction to the actual stock price. The predicted stock price here is for the next day only. 
\\
To get the prediction for the day after that, do this calculation again but take the predicted stock into account.It is possible to do this over and over again for as many days in the future as you would like. Though it is not meaningful for more than a few days because the prediction gets more and more unreliable. For detailed results an the amount of days that are rational, see chapter \ref{results} on page \pageref{results}.

@@

This algorithm is based on analysing the moving average of a stock. It does not make a prediction about the precise prise of a stock the next day. The only thing it does, is giving an advice when to buy a stock and when to sell it. If you follow the advices of this algorithm, it is possible to have a positive outcome even if the trend of a stock is overall going down.
\\
The first thing we had to do was calculating the moving average of a stock. This can be computed as below-
\\
\\
$ MA_{n} =  \frac{C_{D} + C_{D-1} + ... + C_{D-(n-1)}}{n} $
\\
\\
Where C is the closing price of a stock on a specific day D.

Now we need to calculate MA$ _{50} $\footnote{Moving average of the last 50 days} and MA$ _{200} $\footnote{Moving average of the last 200 days} of a specific stock we want to invest in.
\\
If the condition MA$ _{50} >$  MA$ _{200} $ is fulfilled the algorithms gives the advice to buy the stock on this day. After the stock has been bought, the algorithm checks every day if the requirement is still met. As soon as MA$ _{50} <$  MA$ _{200} $ the algorithm recommends to sell the stock. This algorithm seems to be very simple but it is worth noticing, that the outcome was pretty solid in the test we made. For detailed results, see chapter \ref{results} on page  \pageref{results}.

\subsection{Algorithm using neuronal network and machine learning}

It does a prediction based on machine learning and a neuronal network. This algorithm gets a stock prices of the past and tries to learn. Machine learning algorithms learn from and try to make predictions on data. In general these algorithms try to recognize a pattern in the data that is given to them and make predictions based on this patterns.
\\
Our algorithm gets the data of a specific stock and gives a forecast based on the past closing prices of this stock. After we downloaded the data from yahoo finance, we need to normalize the values.
\\
\\
$ normalized(C_{i}) = \frac{C_{i} - C_{min}}{C_{max} - C_{min}} $
\\
\\
With $ C_{min} $ the minimum close price of the stock in the span we examined and $ C_{max} $ as the maximum. The algorithm needs this normalized data to work properly. After we normalized all closing prices of the interval we wanted to investigate, we wrote them into a text file in a specific format. This format is very important, because it is needed for the algorithm to learn properly. 
\\
Every row in the text file got a specific number of normalized values. The first $ n-1 $ values are the ones the algorithm tries to find a specific arithmetic in. The last value in every line is the one, the algorithm tries to predict. After every line, the algorithm reviews its own results and compares it to the real value that is written in the file.
\\
So with every line the algorithm works on, it gets more experience with this kind of prediction. In the last line of our text file, we wrote down the current stock prices until the one of today. The last number of this line is then predicted by algorithm and should be close to the real stock price of the next day.
\\
We tried different values for the number of neurons, lines and values in every line. These changes led to varieties in our results. After we predicted the price for the next day, we did the same algorithm again and took our first prediction as the last day. So we were able to predict more than just one day in advance. Later in the result section, we will show, that going far into the future leads to significant differences in prediction and reality.

\section{Results}
\label{results}

@@

How this algorithm works in detail is described in section \ref{algorithm}. This algorithm is all about predicting the exact price of a stock for the next days. With this algorithm we tried to predict if a stock would go down or up on the next day. We also tried to predict more than just one day in the future. Later in this chapter we will show how good our predictions were on the first and on the nth day in the future. It is understandable that, the more you go into the future, the more unreliable the predictions become. 
\\
To get better and more reliable results, we did not test this method on single stocks. We wanted to get a big amount of predictions and see whether they were good. Because if you only predict one single price it is not really significant, but if you predict many prices and take the average of these predictions you get meaningful results. That is why we took 150 stocks for our algorithm to predict. We tested 3 sets of data with 150 stocks each.
\\
Also a very important thing is the amount of days the algorithm uses to make its prediction. We got the best results using something in the span of 10-15 past days. If you use more than 15 days, it does not really take the latest trend into account and if you use less than 10 days the timespan to get a good prediction is to short.
\\
\begin{center}
\begin{tabularx}{0.49\textwidth}{|X|X|X|X|}
\hline
Days in the future & Set 1 & Set 2 & Set 3\\
\hline
\hline
Day 1. & 0.043 \$ & 0.012 \$ & 0.028 \$\\
\hline
Day 2. & 0.049 \$ & 0.010 \$ & 0.033 \$\\
\hline
Day 3. & 0.059 \$ & 0.013 \$ & 0.041 \$\\
\hline
Day 4. & 0.089 \$ & 0.017 \$ & 0.038 \$\\
\hline
Day 5. & 0.092 \$ & 0.024 \$ & 0.049 \$\\
\hline
Day 6 & 0.126 \$ & 0.035 \$ & 0.062 \$\\
\hline
\hline
Day 15. & 1.740 \$ & 0.092 \$ & 0.107 \$\\
\hline

@@

As described in Chapter \ref{algorithm} this algorithm uses two different MA's to make a good prediction when to buy and when to sell a stock. It is not an algorithm to predict an exact price but rather to make profit without knowing the exact price the stock is going to have the next days.
\\
\\
\begin{figure}[!h]
  \includegraphics{MSFT_MA}
  \caption{Moving Average of MSFT (Microsoft Corporation)}
  \label{fig:MSFT_MA}
\end{figure}

 In Figure \ref{fig:MSFT_MA} you can see the development the stock of Microsoft Corporation\texttrademark made, in the last 11 years. If you look closely, there is also a red and green line in this Figure. The green line is the MA of the last 200 days of every point in this graph. The  red line marks the MA of the last 50 days. Now as described in section \ref{algorithm}, we used these two things to figure out when to buy and when to sell a stock.
\\
We now take a closer look at a single passage of the stock to explain our approach in more detail.
\\
\begin{figure}[!h]
  \includegraphics{MSFT_Zoomed}
  \caption{MSFT stock zoomed}
  \label{fig:MSFT_Z}
\end{figure}

@@

\subsection{3rd algorithm}

@@

\appendix
\section*{list of abbreviations}
MA - Moving average
\\
LTMA - Long term moving average Algorithm
\\
LTA - Latest trend algorithm

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{picture}}]{John Doe}
\blindtext
\end{IEEEbiography}

>>>>

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print - the typesetting of the document will not typically be
% affected with changes in paper size (but the bottom and side margins will).
% Use the testflow package mentioned above to verify correct handling of
% both paper sizes by the user's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
\documentclass[conference]{IEEEtran}
\usepackage[]{blindtext, graphicx}
\usepackage[hidelinks]{hyperref}
\graphicspath{ {Figures/} }
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{color}

@@

There is a huge variety of methods that are used in forecasting stock prices. Many of the algorithms that have been made up just use structured data like tables with historic stock quotes in it. The aim of all these algorithms is to maximize the profit while keeping the risk as low as possible.

Since the stock market is global, there are many relationships between stocks. Some of them are paired with others so if the price of one stock goes down, the other stock will follow this trend.
Stocks are also influenced by numerous factors and these factors also have relationships between each other. This leads us to the thought, that a good stock prediction has to consider more than just past stock prices. However it is still possible to make reasonably good predictions based on just previous prices.

One very important thing that to observe is influence of international newscasts. If the global economy, or even the economy of a single country is experiencing a financial crisis, the stock market will follow that trend. But those trends are most likely to be relatively slow and easy to predict even looking at the stock prices itself. What is more dangerous are catastrophes. After a catastrophe there is often a low point on the stock market. This low point comes so fast, that only algorithms that observe the news have a chance to prevent huge losses.

Another to note is, that we did all our tests on historic data of the stock market. Even if we \textit{bought} a stock at some specific point of time, we did not really buy this stock on the real market. So our tests did not affect the stock market at all. Because buying and selling stocks on the market while testing, would always affect the received results.

@@

In this paper we are going to compare some algorithms and their functionality. The main goal of these algorithms is, to predict a rise or fall in the price before the actual rise or fall happens. It is very important to find some kind of pattern in the stock prices of the past to do a accurate prediction of the future prices. We also like to share some of our experimental results we had with the algorithms. We only used algorithms that do not take global news in account because we wanted to see, how good the algorithms were in an isolated environment analysing past prices only. We decided to do it that way, because the algorithms themselves should work properly and examining news would be an easy addition to these working algorithms.

The main goal of the project was, to find a good working stock forecast algorithm to predict future stock trends. Later in this paper, we will introduce parameters and variables, that give the best results combined with our algorithms.

\subsection{Document overview}
Section \ref{essentials} briefly covers the background information needed for the reader to understand the most important points about our project.

Section \ref{algorithm} describes the algorithms used in more detail.

Chapter \ref{results} is about the test-results received using the algorithms described in chapter \ref{algorithm}. All algorithms were tested with different inputs and parameters.

In section \ref{conclusion} the results are interpreted and we propose how our algorithms can be improved to be successful in the real stock market.

In chapter \ref{futurework} some thing extensions to our algorithms are given, which could be done in future.

@@

We investigated algorithms that bring good results with time series analysis. The computer programs we implemented were specialized  making a good prediction only looking at one single stock and its trend. These algorithms were tested on real stock data we fetched from the yahoo finance API.

We used yahoo finance because it gives easy access to numerous stocks and their historical prices. The API yahoo grants is pretty simple. The only thing you have to do to download the data you need is, preparing a link so it fits your needs. For example the url -

\textit{\url{http://real-chart.finance.yahoo.com/table.csv?s=IBM&a=00&b=2&c=1962&d=04&e=21&f=2015&g=d&ignore=.csv}}
\\
\\
- provides the data in a csv-file. The stock that this link is going to download is IBM (International Business Machines Corporation). It is also possible to adjust the timespan and format of the data. 
This method of getting historical stock market data is pretty easy to implement in various programming languages. It is also possible to download the csv-file to your computer with this link.

After acquiring the necessary data, we started implementing the algorithms that we thought of.

@@

We implemented three different algorithms to predict stock prices and maximise the outcome. All three of them only consider one stock at a time and do not take any global news or trends in account. In this section describes the algorithms used in detail

@@

With this algorithm we wanted to make a precise prediction on the stock market. It analyses the closing prices of a specific stock and gives a forecast how the price will develop. We also wanted the algorithm to be able to predict not only the price for the next day, but for the next few days. In section \ref{results} we will demonstrate that we were able to do a good prediction even 4 days in the future. But it must be said that the more in the future the prediction is, the bigger the gap between the real prices and the prediction.

The algorithm itself uses only the last few days of a stock. The number of days used for this algorithm is not static. The amount of days that give the best result may vary for different stocks. Nevertheless if the number of days stays within a specific range, the results are pretty accurate.
\\
\\
\[ M = \sum_{n=0}^D C_{n+1} - C_{n} \]
\[ prediction = \frac{M}{D}\]
\[ stock_{predicted} = stock_{today} + prediction\]
\\
Where D is the number of days the prediction is based on and C is the closing price of the nth day. After you calculated the prediction, the only thing that is left to do, is adding the prediction to the actual stock price. The predicted stock price here is for the next day only.

To get the prediction for the day after that, do this calculation again but take the predicted stock into account.It is possible to do this over and over again for as many days in the future as you would like. Though it is not meaningful for more than a few days because the prediction gets more and more unreliable. For detailed results an the amount of days that are rational, see chapter \ref{results} on page 
\pageref{results}.

@@

This algorithm is based on analysing the moving average of a stock. It does not make a prediction about the precise prise of a stock the next day. The only thing it does, is giving an advice when to buy a stock and when to sell it. If you follow the advices of this algorithm, it is possible to have a positive outcome even if the trend of a stock is overall going down.

The first thing we had to do was calculating the moving average of a stock. This can be computed as below-
\\
\\
$ MA_{n} =  \frac{C_{D} + C_{D-1} + ... + C_{D-(n-1)}}{n} $
\\
\\
Where C is the closing price of a stock on a specific day D.

Now we need to calculate MA$ _{50} $\footnote{Moving average of the last 50 days} and MA$ _{200} $\footnote{Moving average of the last 200 days} of a specific stock we want to invest in.

If the condition MA$ _{50} >$  MA$ _{200} $ is fulfilled the algorithms gives the advice to buy the stock on this day. After the stock has been bought, the algorithm checks every day if the requirement is still met. As soon as MA$ _{50} <$  MA$ _{200} $ the algorithm recommends to sell the stock. This algorithm seems to be very simple but it is worth noticing, that the outcome was pretty solid in the test we made. For detailed results, see chapter \ref{results} on page  \pageref{results}.

\subsection{Algorithm using supervised learning techniques}

In forecasting future stock trends it is common to not use the stock prices itself. Instead other values based on the stock prices are first calculated and then used as predictor variables for the learning process. One possible value is the daily return which is:
\\
\\
$Return_i = \frac{Close_i - Close_{i-1}}{Close_{i-1}}$
\\
\\
For this algorithm three different techniques are used to learn the model of the stock trends. Those are Logistic Regression, Linear Discriminant Analysis and Quadratic Discriminant Analysis. To test them the returns of the past two days are used as predictor variables.

Through Logistic Regression the probability of a following day being categorized as "Up" or "Down" is modelled. The function used for logistic regression is the following:
\\
\\
$p(Y = U | L_1, L_2) = \frac{e^{\beta_0 + \beta_1 L_1 + \beta_2 L_2}}{1 + e^{\beta_0 + \beta_1 L_1 + \beta_2 L_2}}$
\\
\\
Where $p(Y = U)$ is the probability that the next day's stock price is increasing. $L_1$ and $L_2$ are the lagged returns of the past two days.

Linear and Quadratic Discriminant Analysis both treat their predictor variables independently at first and combine them later using Bayes' theorem.
\section{Results}
\label{results}

@@

How this algorithm works in detail is described in section \ref{algorithm}. This algorithm is all about predicting the exact price of a stock for the next days. With this algorithm we tried to predict if a stock would go down or up on the next day. We also tried to predict more than just one day in the future. Later in this chapter we will show how good our predictions were on the first and on the nth day in the future. It is understandable that, the more you go into the future, the more unreliable the predictions become.

To get better and more reliable results, we did not test this method on single stocks. We wanted to get a big amount of predictions and see whether they were good. Because if you only predict one single price it is not really significant, but if you predict many prices and take the average of these predictions you get meaningful results. That is why we took 150 stocks for our algorithm to predict. We tested 3 sets of data with 150 stocks each.

Also a very important thing is the amount of days the algorithm uses to make its prediction. We got the best results using something in the span of 10-15 past days. If you use more than 15 days, it does not really take the latest trend into account and if you use less than 10 days the timespan to get a good prediction is to short.

\begin{center}
\begin{tabularx}{0.49\textwidth}{|X|X|X|X|}
\hline
Days in the future & Set 1 & Set 2 & Set 3\\
\hline
\hline
Day 1. & 0.043 \$ & 0.012 \$ & 0.028 \$\\
\hline
Day 2. & 0.049 \$ & 0.010 \$ & 0.033 \$\\
\hline
Day 3. & 0.059 \$ & 0.013 \$ & 0.041 \$\\
\hline
Day 4. & 0.089 \$ & 0.017 \$ & 0.038 \$\\
\hline
Day 5. & 0.092 \$ & 0.024 \$ & 0.049 \$\\
\hline
Day 6 & 0.126 \$ & 0.035 \$ & 0.062 \$\\
\hline
\hline
Day 15. & 1.740 \$ & 0.092 \$ & 0.107 \$\\
\hline

@@

As described in Chapter \ref{algorithm} this algorithm uses two different MA's to make a good prediction when to buy and when to sell a stock. It is not an algorithm to predict an exact price but rather to make profit without knowing the exact price the stock is going to have the next days.

\begin{figure}[!h]
  \includegraphics{MSFT_MA}
  \caption{Moving Average of MSFT (Microsoft Corporation)}
  \label{fig:MSFT_MA}
\end{figure}

 In Figure \ref{fig:MSFT_MA} you can see the development the stock of Microsoft Corporation\texttrademark made, in the last 11 years. If you look closely, there is also a red and green line in this Figure. The green line is the MA of the last 200 days of every point in this graph. The  red line marks the MA of the last 50 days. Now as described in section \ref{algorithm}, we used these two things to figure out when to buy and when to sell a stock.

We now take a closer look at a single passage of the stock to explain our approach in more detail.

\begin{figure}[!h]
  \includegraphics{MSFT_Zoomed}
  \caption{MSFT stock zoomed}
  \label{fig:MSFT_Z}
\end{figure}

@@

\subsection{Algorithm using supervised learning techniques}

The algorithms were implemented in python utilizing the scikit-learn library.

The following table shows the results of applying different machine learning techniques to stock data. All used methods achieve similar results. The accuracy of the prediction is barely higher than the probability of a coin flip. But none of the cases do have a prediction accuracy below 50\%. The algorithm, as it is, should not be used without further improvements and refinement.

\begin{center}
\begin{tabularx}{0.5\textwidth}{|X|X|X|X|X|X|}
\hline
{} &     
\^{}AXJO\footnote{Australia ASX-200} &     
\^{}FCHI\footnote{Paris CAC 40} &     
\^{}FTSE\footnote{London FTSE-100} &    
\^{}GDAXI\footnote{Frankfurt DAX} & 
\^{}IXIC\footnote{NASDAQ Composite}\\
\hline
LDA &  0.501 &  0.541 &  0.527 &  0.555 & 0.571\\
LR  &  0.501 &  0.541 &  0.527 &  0.555 & 0.571\\
QDA &  0.541 &  0.513 &  0.519 &  0.539 & 0.575\\
\hline
\end{tabularx}
\\[5pt]
Table 3. Results of Supervised Learning Algorithms
\end{center}

@@

\appendix
\section*{list of abbreviations}
MA - Moving average
\\
LTMA - Long term moving average Algorithm
\\
LTA - Latest trend algorithm
\\
LR - Logistic Regression
\\
LDA - Linear Discriminant Analysis
\\
QDA - Quadratic Discriminant Analysis

@@

%\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{picture}}]{John Doe}
%\blindtext
%\end{IEEEbiography}
